---
title: "PA1_template for Reproducible Research - Assignment 1"
author: "Adrian Johnston"
date: "18 October 2015"
output: html_document
---
###Loading and preprocessing the data
```{r}
#read activity data file which has been unzipped into the directory below
#would be cool to find how to read directly from the original zip in github
setwd("/home/adrian/Rep_Research")
df.activity <- read.csv("activity.csv", header=TRUE, sep=",")

```

###What is mean total number of steps taken per day?
```{r}
#Aggregate activity by date, finding sum of steps per day (ignoring NA as instructed)
df.activity_sum <-setNames(aggregate(x=df.activity$steps, by=list(df.activity$date), FUN="sum", na.rm=TRUE), c("Date", "Total_Steps"))
hist(x=df.activity_sum$Total_Steps, xlab = "Total Steps Per Day", main = "Activity Data Histogram")
```

```{r}
#Calculate mean of Total Steps per day, ignoring NA values
mean_steps<-mean(df.activity_sum$Total_Steps, na.rm=TRUE)
#Calculate median of Total Steps per day, ignoring NA values
median_steps<-median(df.activity_sum$Total_Steps, na.rm=TRUE)
```

The mean of the total steps taken per day in this data set is `r format(mean_steps, digits = 5)`.

The median of the total steps taken per day in this data set is `r format(median_steps, digits = 5)`.

###What is the average daily activity pattern?
```{r}
#Aggregate activity by date, finding sum of steps per day (ignoring NA as instructed)
df.activity_interval <-setNames(aggregate(x=df.activity$steps, by=list(df.activity$interval), FUN="mean", na.rm=TRUE), c("interval", "Average_Steps"))

plot(df.activity_interval, type="l", main = "Average Daily Activity Pattern")

```

The 5-minute interval which, on average across all the days in the dataset, contains the maximum number of steps, is `r df.activity_interval[(df.activity_interval$Average_Steps==
                       max(df.activity_interval$Average_Steps)),"interval"]`

###Imputing missing values

The total number of missing values in the dataset (where step count appears as NA) is 
`r sum(is.na(df.activity$steps))`

Missing values are replaced with the mean steps for that interval.

``` {r}
#The adjusted data frame is df.activity_adj
df.activity_adj<-merge(df.activity, df.activity_interval)

df.activity_adj$steps <- ifelse(is.na(df.activity_adj$steps), df.activity_adj$Average_Steps,                              df.activity_adj$steps) 

df.activity_adj_sum <-setNames(aggregate(x=df.activity_adj$steps, by=list(df.activity_adj$date), FUN="sum"), c("Date", "Total_Steps"))
hist(x=df.activity_adj_sum$Total_Steps, xlab = "Total Steps Per Day", main = "Activity Data Histogram", sub="(Missing Values Imputed)")

```


```{r}
#Calculate mean of Total Steps per day, ignoring NA values
mean_steps_adj<-mean(df.activity_adj_sum$Total_Steps)
#Calculate median of Total Steps per day, ignoring NA values
median_steps_adj<-median(df.activity_adj_sum$Total_Steps)

Adjustment_Impact_Mean <- mean_steps - mean_steps_adj
Adjustment_Impact_Median <- median_steps - median_steps_adj
```

The adjusted mean of the total steps taken per day in this data set is `r format(mean_steps_adj, digits = 5)`.

The adjusted median of the total steps taken per day in this data set is `r format(median_steps_adj, digits = 5)`.

The impact of the adjustment on the mean equals to `r format(Adjustment_Impact_Mean, digits = 5)` steps.
(Original Mean - Adjusted Mean)

The impact of the adjustment on the mean equals to `r format(Adjustment_Impact_Median, digits = 5)` steps.
(Original Median - Adjusted Median)

###Are there differences in activity patterns between weekdays and weekends?

```{r}
df.activity_adj$Weekday<-weekdays(as.Date(df.activity_adj$date))
df.activity_adj$Day_Type <-as.factor(ifelse(weekdays(as.Date(df.activity_adj$date)) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
```

Time series plot showing average number of steps taken, averaged across all weekday days or weekend days.

```{r}
#Prepare panel plot
library(lattice)
df.activity_adj_sum_by_Day_Type<-setNames(aggregate(x=df.activity_adj$steps,
by=list(df.activity_adj$interval, df.activity_adj$Day_Type), FUN="mean"), c("Interval", "Day_Type", "Average_Steps"))

#attach(df.activity_adj_sum_by_Day_Type)

xyplot(df.activity_adj_sum_by_Day_Type$Average_Steps ~          df.activity_adj_sum_by_Day_Type$Interval|df.activity_adj_sum_by_Day_Type$Day_Type, 
type="l", ylab = "Number of Steps", xlab = "Interval", layout = c(1, 2))




